---
title: "An Introduction to Statistical Data Analysis in R^[These slides are not intended to be self-contained and comprehensive, but just aim to provide some of the workshop's content. Much more will be provided in the workshop itself.]"
author: |
  | Mark Andrews
  | Psychology Department, Nottingham Trent University
fontsize: 10pt
output:
 beamer_presentation:
  keep_tex: false
  fonttheme: "serif"
  includes:
   in_header: slides_preamble.tex
---



```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = T, warning = FALSE, message = FALSE)
options(scipen=9)
set.seed(42)

library(tidyverse)

available_packages <- available.packages(repos = "http://cran.us.r-project.org") %>% as_tibble()

open_licences <- str_detect(available_packages$License, 
                            regex('gpl|bsd|mit|apache|cc|GNU|artistic', ignore_case = T)) %>% 
  mean()
```



# Linear regression

* Predict `ACT` as a linear function of `education` in the `sat_act` data frame.
```{r}
sat_act <- read_csv('../data/sat_act.csv')
M <- lm(ACT ~ education, data=sat_act)
summary(M)
```

# Predictions in linear regression

* On the basis of our fitted model ```M```, we can make predictions about possible values of the predictor variable. 
```{r}
hypothetical_data <- data.frame(education = c(1, 2, 5, 10, 15))
predict(M, newdata=hypothetical_data)
```


# Multiple linear regression

* We can add as many predictor variables as we like.
```{r}
M <- lm(ACT ~ education + age + gender, data=sat_act)
summary(M)
```

# Collinearity

* We'll evaluate multicollinerity using Variance Inflation Factor (VIF):
```{r}
library(car)
vif(M)
```

# General linear models

* We can use predictors that categorical as well as continuous in our model. 
* Here, we investigate how the post treatment weight of a patient differs from their pre treatment weight, for three different types of therapy (control, CBT, family therapy).

```{r}
anorexia <- read_csv('../data/anorexia.csv')
```

# General linear models (continued)

* First, we'll visualize the data.
```{r, out.width='0.75\\textwidth',out.extra='keepaspectratio', fig.align='center'}
ggplot(anorexia,
       aes(x = Prewt, y = Postwt, col=Treat)) +
  geom_point() +
  stat_smooth(method='lm', se=F) +
  theme_classic()
```

# General linear models (continued)

* Here, we do a *varying intercept*, which is also known as an *ANCOVA*:
```{r}
M <- lm(Postwt ~ Prewt + Treat, data=anorexia)
summary(M)
```

# General linear models (continued)

* We cam also do a *varying slopes and varying intercepts* model. This is a type of interaction model:
```{r}
M_interaction <- lm(Postwt ~ Prewt * Treat, data=anorexia)
summary(M_interaction)
```

# Model evaluation

* We can compare any two linear models using the generic ```anova``` function. 
* Here, we'll use this to test whether the varying slopes and intercepts model is a better fit to the data than the just varying intercepts model:

```{r}
anova(M, M_interaction)
```

# One-way Anova

* We can use `aov` for one-way (and other) Anova.
```{r}
data(PlantGrowth)
M <- aov(weight ~ group, data=PlantGrowth)
summary(M)
```

# Multiple comparisons

* We can do Tukey's range test to perform multiple comparisons:
```{r}
TukeyHSD(M)
```

# One-way Anova (alternative)

* Note that we can also we can do Anova using `lm()`:
```{r}
M <- lm(weight ~ group, data=PlantGrowth)
anova(M)
```


# Two-way anova

```{r, out.width='0.75\\textwidth',out.extra='keepaspectratio', fig.align='center'}
data("ToothGrowth")

ggplot(ToothGrowth,
       aes(x = factor(dose), y = len, col = supp)) +
  geom_boxplot() +
  theme_classic()
```


# Two-way (factorial) anova

```{r}
M <- aov(len ~ supp*dose, data=ToothGrowth)
summary(M)
```

# One-way repeated measures Anova

```{r}
recall_data <- read_csv('../data/recall_data.csv')

M <- aov(Recall ~ Valence + Error(Subject/Valence), data=recall_data)
summary(M)
```

# One-way repeated measures Anova (continued)

* Multiple comparisons, with Bonferroni correction
```{r}
with(recall_data, 
     pairwise.t.test(x=Recall, g=Valence), 
     p.adjust.methods='bonferroni', 
     paired=T)
```

# Twoway repeated measures Anova
```{r}
recall_data2 <- read_csv('../data/recall_data2.csv')
M <- aov(Recall ~ Valence*Task + Error(Subject/(Task*Valence)), 
         data=recall_data2)
summary(M)
```


# Multilevel models

* The repeated measures anova above can be done, and I think *should* be done, using multilevel models too.

```{r}
library(lme4)
M <- lmer(Recall ~ Valence*Task + (1|Subject),
          data=recall_data2)
```



